---
title: 'One Hundred and Fifty Years of Warming on Caribbean Coral Reefs'
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r knit options, include=FALSE}

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE)

```

```{r load libraries}

### Used packages that need to be installed to run code:
# install.packages("sf", "ncdf4", "ggrepel", "raster", "viridis", "maptools", "rgdal", "ggplot2", "cowplot", "tidyverse", "rnaturalearth", "xts", "RColorBrewer", "tidyr", "dplyr", "binr", "fields", "gganimate", "sp", "mgcv", "plotly", "grDevices", "repmis", "foreach", "kableExtra")

# library(devtools)
# devtools::install_github("NCAR/fields")
# devtools::install_github("thomasp85/transformr") # need this for the GIFs of figures
# devtools::install_github("paleolimbot/ggspatial")

library(sf)
library(ncdf4)
library(ggrepel)
library(raster) 
library(viridis)
library(maptools)
library(rgdal)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(rnaturalearth)
library(xts)
library(RColorBrewer)
library(tidyr)
library(dplyr)
library(binr)
library(fields)
library(gganimate)
library(sp)
library(mgcv)
library(foreach)
library(plotly)
library(kableExtra)
library(grDevices)
library(repmis)

```

```{r other setup}

## set the current date
date <- Sys.Date() 


## spatial subsetting for Caribbean region
Xmin <- -100
Xmax <- -55
Ymin <- 0
Ymax <- 40


## set the projection
proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84"


## set the binning of p-values for plotting
bins <- c(-Inf, 0.0001, 0.001, 0.01, 0.05, 0.1, Inf)
bin_names <- c("< 0.0001", "0.0001 - 0.001", "0.001 - 0.01", "0.01 - 0.05", "0.05 - 0.1", "> 0.1")


## Create base map for SST maps
map_base <- ggplot2::fortify(maps::map(fill = TRUE, plot = FALSE)) %>% 
  dplyr::rename(lon = long)


## Make a shapefile of lat/lon bounds of MHW map
# make dataframe of lat/lon
bound_df <- data.frame("lon" = c(262.125, 285.125, 285.125, 302.125, 302.125, 285.125, 285.125, 276.125, 276.125, 271.125, 269.625, 262.125,262.125),
                       "lat" = c(31, 31, 25.125, 25.125, 9.125, 9.125, 7.375, 7.375, 15.875, 15.875, 17.875, 17.875, 31))
bound_df$lon <- bound_df$lon - 360 # convert lon back into -180 to 180
bound_matrix <- data.matrix(bound_df) # Transform from dataframe to a matrix
bound_poly <- Polygon(bound_matrix) # transform matrix to polygon
bound_SP <- SpatialPolygons(list(Polygons(list(bound_poly), ID = "a")), proj4string=CRS(proj)) # convert polygon to SpatialPolygon object
bound_SP_df <- as(bound_SP, "SpatialPolygonsDataFrame") # convert to SpatialPolygonsDataFrame

```

```{r setting non-git file paths}

### Since the shape and netCDF files are too large to store on GitHub, those need to be stored locally.
## Dropbox links are included to download the corresponding files. Just update the local path to each.
## The paths for the files stored on Dropbox are below:

## gshhs_f.b: coastline map shapefile
# Dropbox link: 
gshhs_f_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/gshhs_f.b"

## WCMC008_CoralReef2018_Py_v4.shp: global coral reef shapefiles
# Dropbox link: 
WCMC008_CoralReef2018_Py_v4_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/14_001_WCMC008_CoralReefs2018_v4/01_Data/WCMC008_CoralReef2018_Py_v4.shp"

## meow_ecos.shp: global ecoregion shapefiles
# Web link: https://www.worldwildlife.org/publications/marine-ecoregions-of-the-world-a-bioregionalization-of-coastal-and-shelf-areas
ecoregion_shape_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/MEOW/meow_ecos.shp"

## HadISST_sst.nc: 
# Dropbox link: 
HadISST_sst_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/HadISST_sst.nc"

## pathfinder_combined_monthly_data.nc: 
# Dropbox link: 
pathfinder_combined_monthly_data_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/pathfinder_combined_monthly_data.nc"

## Pathfinder_monthly_edit.nc: 
# Dropbox link:
Pathfinder_monthly_edit_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/PathfinderSST_monthly_edit.nc"

## OISST_MHW_carib.Rda: 
# Dropbox link: 
MHW_path <- "/Users/colleen/Dropbox/Git/Data_Not_On_Git/CaribbeanSST_data/OISST_MHW_carib.Rda"

```

<br/>

#### Colleen B. Bove^1,2^*, Laura Mudge^1^, and John F. Bruno^1^

^1^ The Department of Biology, The University of North Carolina at Chapel Hill, Chapel Hill, North Carolina, 27599-3280 USA. 

^2^ The Department of Biology, Boston University, Boston, Massachusetts, 02215 USA. 

* Corresponding author: colleenbove@gmail.com

<br/>

#### **Abstract:**

*Abstract text will go here*

<br/>

##### **Citation:**

*Citation will go here*


<br/>

## Caribbean Ecoregion Coral Reefs {.tabset}

### <span style="color: #0072B2;">**Figure 1**</span>

```{r read in land shapefile, include=FALSE}

## read in the land shapefile with slightly expanded lat/lon limits
if (!rgeosStatus()) gpclibPermit()
gshhs.f.b <- gshhs_f_path
land <- getRgshhsMap(gshhs.f.b, xlim = c(-105, -40), ylim = c(-10, 40)) %>%
  fortify()

```

```{r create Caribbean reef shapefile, eval=FALSE, include=FALSE}

## read in the global coral reef shapefile 
shape <- shapefile(WCMC008_CoralReef2018_Py_v4_path)
reefs_cropped <- crop(shape, extent(Xmin, Xmax, Ymin, Ymax)) # crop for the Caribbean region only
shapefile(reefs_cropped, "data/ReefData/Caribbean_reefs.shp") # save the new Caribbean-specific reef shapefile

```

```{r assign reefs to ecoregions, message=FALSE, warning=FALSE, paged.print=FALSE}

## Read in the reef points csv and clipped Caribbean ecoregion shapefile
reef_gps <- read.csv("data/ReefData/CaribbeanReef_points.csv") # read in FULL reef locations csv
ecoreg <- sf::read_sf("data/EcoRegions/Caribbean_ecoregions.shp") # this is used for ecoregion name list (plotting ecoregion bounds)
ecoregions2 <- readOGR('data/EcoRegions', 'Caribbean_ecoregions') # this is used for assigning ecoregion in the forloop below


## Subset Caribbean shapefile for the TNA (The North Atlantic) province (code 12) with Bermuda removed and only the N. GoM within the WTNA province (codes 6)
carib_ecoreg <- ecoreg %>% 
  filter(PROV_CODE == 12 & ECOREGION != "Bermuda" | PROV_CODE == 6 & ECOREGION == "Northern Gulf of Mexico") 

# create unique reef id
reef_points <- reef_gps %>%
  mutate(reef_id = as.numeric(interaction(lat, lon, drop = TRUE))) %>%
  filter()

# transform points into SpatialPointsDataFrames and modify projection to be the same
coordinates(reef_points) <- c("lon", "lat")  
projection(reef_points) <- proj # change the projection of the reef_points dataframe
projection(ecoregions2) <- proj # change the projection of the ecoregions2 polygopn dataframe


## Assign reef locations to ecoregions (using forloop)

eco_list <- carib_ecoreg$ECOREGION # create a list of the ecoregions (used in the forloop)

# make a blank dataframe to be populated from the forloop
compiled_eco_df <- data.frame('lat' = integer(),
                              'lon' = integer(),
                              'reef_id' = factor(),
                              'ecoregion' = factor())

## Create a forloop to assign reef loactions to each ecoregion and compile into single dataframe
for(e in 1:length(eco_list)) {
  ecoregion_name <- eco_list[[e]] # pull the name of ecoregion
  subset_ecoreg <- ecoregions2[ecoregions2@data$ECOREGION == ecoregion_name, ] # subset for only the desired ecoregion
  projection(subset_ecoreg) <- proj # modify the lat/lon projection to match reef points
  ecoregion_reefs <- reef_points[subset_ecoreg, ] # clip the reef points by the ecoregion bounds
  ecoregion_df <- as.data.frame(ecoregion_reefs) # transform reef points (spatial) to dataframe
  ecoregion_df$reef_id <- factor(ecoregion_df$reef_id) # make reef ID into a factor
  ecoregion_df$ecoregion <- rep(ecoregion_name, length(ecoregion_df$reef_id)) # add ecoregion name to the data
  compiled_eco_df <- rbind(compiled_eco_df, ecoregion_df) # add the most recent layer to the full dataframe
}


# Rename "Northern GoM" and "Northern GoM" to just GoM (these will be considered the same ecoregion from here on)
compiled_eco_df$ecoregion <- gsub("Northern G", "G", compiled_eco_df$ecoregion)
compiled_eco_df$ecoregion <- gsub("Southern G", "G", compiled_eco_df$ecoregion)

```

```{r reefs by ecoregion map}

### Plot the map (reefs are coloured by ecoregion with # of reefs beside region name)

## Update ecoregion labels to include number of reefs within each 
# create dataframe of just ecoregion and n
lab_df <- compiled_eco_df %>% 
  group_by(ecoregion) %>% 
  summarize(n = n())

# create new labels to display 'ECOREGION (n=NUMBER_REEFS)'
labels <- paste(lab_df$ecoregion, " (n=", lab_df$n, ")", sep = "")

# Name existing ecoregion labels based on new labels for plot 
eco_labs <- c("Bahamian" = labels[1],
              "Eastern Caribbean" = labels[2],
              "Floridian" = labels[3],
              "Southwestern Caribbean" = labels[7],
              "Western Caribbean" = labels[8],
              "Southern Caribbean" = labels[6],
              "Gulf of Mexico" = labels[5],
              "Greater Antilles" = labels[4])


## set colour palette and read in world shapefile
world <- ne_countries(scale = "medium", returnclass = "sf") # world shapefile
pal <- c("#FF948C", "#E48732", "#387560", "#7993CF", "#6E4096", "#5898AF", "#D9B03B", "#B8321D") # colour palette for ecoregions
pal2 <- c("#FF948C", "#E48732", "#387560", "#7993CF", "#6E4096", "#5898AF", "#6E4096", "#D9B03B", "#B8321D") # ecoregion bounds colours (see note below)


## Map plot
# Note: to display the ecoregion bounds, unhash the two lines to plot the shapefiles over the map

ggplot() + 
  geom_point(data = compiled_eco_df, aes(x = lon, y = lat, colour = ecoregion), size = 0.7) +
  geom_sf(data = world) +
  #geom_sf(data = carib_ecoreg, colour = "black", aes(fill = ECOREGION), alpha = 0.1) + # for showing ecoregion bounds
  #scale_fill_manual("Ecoregion", values = pal2, labels = eco_labs) + # for showing ecoregion bounds (colours)
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), legend.text=element_text(size=12), legend.title=element_text(size=14), axis.text = element_text(size = 10)) +
  guides(colour = guide_legend(override.aes = list(size = 4))) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33)) +
  scale_colour_manual("Ecoregion", values = pal, labels = eco_labs) +
  ylab("") +
  xlab("") +
  ggsave("figures/Manuscript_figs/Fig1_ReefMap_map.pdf", width = 8, height = 3.5)

```

**Caribbean coral reef site locations and ecoregion designation.** The colour of each reef represents the designated ecoregion (Spalding et al. 2010) and n denotes the number of unique reefs within that ecoregion.

<br/>


## SST Extraction and Analysis  {.tabset}

### <span style="color: #0072B2;">**Figure 2**</span>

```{r HadISST raster extraction}

## read in the HadISST raster
had_data <- HadISST_sst_path # this is from the path defined at the start
data_brick <- brick(had_data) # read SST data as rasterbrick
data_brick <- crop(data_brick, extent(Xmin, Xmax, Ymin, Ymax)) # clip rasterbrick for Caribbean region only (see 'spatial subsetting' above)


ncdf_file <- nc_open(had_data) # open up the netCDF

nc_lats <- ncdf_file$dim$latitude$vals #other way of looking into the data structure; extracting lats (HadISST)
nc_longs <- ncdf_file$dim$lon$vals #longs here (HadISST)
nc_longs[nc_longs>180] <- nc_longs[nc_longs>180]-360 # convert lat/lon

#time handling
raw_times <- ncdf_file$dim$time$vals #times are seconds after the "start of time"
startoftime <- substr(ncdf_file$dim$time$units,1,19) #HadISST
nc_times <- as.POSIXct(gsub("days since ","",startoftime), tz="UTC", format='%Y-%m-%d') + (raw_times*60*60*24) #HadISST
nc_times <- nc_times+3600*3 #to center readings at noon - just cosmetics
nc_close(ncdf_file) #close the file

#temporal subsetting  
start_date <- "1871/01/01"
end_date <- "2020/12/31"
start_date <- as.POSIXct(start_date,tz="UTC")
end_date <- as.POSIXct(end_date,tz="UTC")

##restrict temporally - here we are not restricting temporally so not data is being removed here
Tstart_index <- min(which(difftime(nc_times,start_date)>0))
#Tend_index <- min(which(difftime(nc_times,end_date)>0))
data_brick <- data_brick[[Tstart_index:length(nc_times)]]
time <- nc_times[Tstart_index:length(nc_times)]

save(time, nc_times, file = "data/HadISST/HadISST_times.Rdata") # save the time variables for IDs for running scripts remotely


### subsetting the HadISST raster for 1981 - 2019 timespan for comparison with Pathfinder
start_dateb <- "1981-08-16" #"1990-01-16" 
start_dateb <- as.POSIXct(start_dateb, tz="UTC")
Tstart_indexb <- min(which(difftime(nc_times, start_dateb)>0))

data_brickb <- brick(had_data)
data_brickb <- crop(data_brickb, extent(Xmin, Xmax, Ymin, Ymax))
data_brickb <- data_brickb[[Tstart_indexb:length(nc_times)]]
timeb <- nc_times[Tstart_indexb:length(nc_times)]

```

```{r HadISST lm map df creation}

## simple lm function applied to the full HadISST raster brick
lm.fun <- function(x, time)
{
  if(any(is.na(x))){NA}
  else{
    lm(x~time)$coefficients[2]*60*60*24 # slope here is degrees/day
  }
}
raster_slope <- calc(data_brick, function(x)lm.fun(x, time = time))

# convert the raster to a df for plotting
raster_slope <- raster_slope*(365.25*10) # this is currently C per decade
had_slope <- as.data.frame(as(raster_slope, "SpatialPixelsDataFrame")) # convert to a raster to dataframe
colnames(had_slope) <- c("sst", "x", "y")


########## Same idea, this time only being used on the 1981 - 2019 subset

# apply the function
raster_slopeb <- calc(data_brickb, function(x)lm.fun(x, time = timeb))

# convert the raster to a dataframe for plotting
raster_slope_sub <- raster_slopeb*(365.25*10) # this is currently C per decade
had_slope_sub <- as.data.frame(as(raster_slope_sub, "SpatialPixelsDataFrame")) # convert to a raster to dataframe
colnames(had_slope_sub) <- c("sst", "x", "y")

# --------- *Plotting occurs down at the bottom of the markdown file* --------- #

```

```{r HadISST pvalue map df creation}

## function to pull the p value per pixel of the simple lm applied to the full HadISST raster brick 
pval.fun <- function(x, time)
{
  if(any(is.na(x))){NA}
  else{
    summary(lm(x ~ time))$coefficients[4] # pulls the p value
  }
}
raster_pval <- calc(data_brick, function(x)pval.fun(x, time = time))

# convert the raster to a df for plotting
had_pavl <- as.data.frame(as(raster_pval, "SpatialPixelsDataFrame")) # convert to a raster to dataframe
colnames(had_pavl) <- c("pval", "x", "y")
had_pavl$bins <- cut(had_pavl$pval, breaks = bins, labels = bin_names) # bins the pvales into 6 unique bins for plotting


########## Same idea, this time only being used on the 1981 - 2019 subset

# apply function to pull p value from each grid
raster_pvalb <- calc(data_brickb, function(x)pval.fun(x, time = timeb))

# convert the raster to a df for plotting
had_pval_sub <- as.data.frame(as(raster_pvalb, "SpatialPixelsDataFrame")) # convert to a raster to dataframe
colnames(had_pval_sub) <- c("pval", "x", "y")
had_pval_sub$bins <- cut(had_pval_sub$pval, breaks = bins, labels = bin_names) # bins the pvales into 6 unique bins for plotting

# --------- *Plotting occurs down at the bottom of the markdown file* --------- #

```

```{r HadISST reef SST extraction, eval=FALSE, include=FALSE}

## read in FULL reef locations csv
reef_gps <- read.csv("data/ReefData/CaribbeanReef_points.csv")

## convert sampling points to spatial points and applying projection to points
LongLat <- cbind(reef_gps$lon, reef_gps$lat) 
ReefPTS <- SpatialPoints(LongLat)
proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84"
projection(ReefPTS) <- proj

## extract data from the sst brick with the sampling points shifted
had_sst <- raster::extract(data_brick, ReefPTS)

## convert the data matrix to a final dataframe
had_sst_xts <- xts(t(had_sst), time) # t is used to transpose the matrix because xts assumes dates are along the lines, not columns
had_sst_df <- data.frame(date=index(t(had_sst_xts)), coredata(t(had_sst_xts))) # convert xts object to dataframe
had_sst_df$reef <- factor(seq(1, length(had_sst_df$date), 1)) # ID the columns to identify unique reefs
had_sst_df$date <- NULL # remove this column since it is repetitive 
had_sst_long <- gather(had_sst_df, date, sst, X1871.01.16.15.00.00:X2020.12.16.15.00.00) # convert from wide to long format

## update the dataframe for saving
had_gps <- had_sst_long %>% separate(date, c("year", "month", "day", "hour", "min", "sec")) # create year, month, and day column columns
had_gps$date <- paste(had_gps$year, had_gps$month, had_gps$day, sep = ".") # creates a full date column (XYear.Month.Day)
had_gps <- had_gps[, -c(2:7)] # remove columns for hour, minute, and seconds (we won't use these)
had_gps <- spread(had_gps, reef, sst)

## save the final dataframe (.csv or Rdata)
#write.csv(had_gps, file = "data/HadISST/HadISST_reefs_SST_update.csv") # to save file as .csv
save(had_gps, file = "data/HadISST/HadISST_reefs_SST_update.Rdata") # to save file as Rdata (I saved as Rdata since it is large)

```

```{r HadISST monthly reef SST dataframe construction}

load("data/HadISST/HadISST_reefs_SST_update.Rdata")

had_gps <- had_gps # only need to use this to make the UPDATED reef SST dataframe compatible below

## create a new dataframe of simple stats per time
had_full <- data.frame("date" = had_gps[,1],
                      "mean_temp" = rowMeans(had_gps[,-1], na.rm = TRUE),
                      "max_temp" = apply(had_gps[,-1], 1, FUN=max, na.rm = TRUE),
                      "min_temp" = apply(had_gps[,-1], 1, FUN=min, na.rm = TRUE),
                      "n" = length(had_gps[,-1]))

had_full <- had_full %>% separate(date, c("year", "month","day"), convert = TRUE) # create year, month, and day column columns
had_full$year <- as.numeric(gsub("\\X", "", had_full$year)) # convert the year column to numerics and remove 'X'
had_full$month2 <- factor(had_full$month) # make a month as a factor column for plotting

```

```{r HadISST monthly SST per year}

#### Data visualization of monthly SST per year #### 

# calculate summary statistic per month across years
HADI_month_avg <- had_full %>% 
  group_by(month) %>% 
  summarize(mean = mean(mean_temp, na.rm = TRUE),
            sd = sd(mean_temp, na.rm = TRUE),
            n = n()) %>% 
  mutate(se = sd / sqrt(n),
         lowerci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upperci = mean + qt(1 - (0.05 / 2), n - 1) * se)

# make year integer for better plotting
had_full$year <- as.integer(had_full$year)

## create plots of monthly HadISST per year
had_month_plot <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 14)) +
  theme(legend.position = c(0.065, 0.85), legend.key.size = unit(0.3, "cm"), legend.background = element_rect(fill = NA), plot.title = element_text(size = 12)) + # legend theme
  guides(colour = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
  geom_line(data = had_full, aes(x = month, y = mean_temp, group = year, colour = year), alpha = 0.6, size = 0.4) +
  scale_colour_gradient2("", low = "#2166ac", mid = "#d1e5f0", high = "#d6604d", midpoint = mean(had_full$year), space="Lab", breaks = seq(1870, 2020, 50)) +
  #geom_ribbon(data = HADI_month_avg, aes(x = month, ymin = lowerci, ymax = upperci), colour = "black", alpha = 0.6, linetype = 2) + 
  #geom_line(data = HADI_month_avg, aes(x = month, y = mean), colour = "black", group = 1, size = 1.5) +
  scale_x_continuous(name = "Month", breaks = c(1,2,3,4,5,6,7,8,9,10,11,12), labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  scale_y_continuous(breaks = seq(24, 30, 1)) +
  ylab('SST (°C)') +
  ggtitle("HadISST (1871 - 2020)") + 
  theme(plot.title = element_text(size = 13))

```

```{r HadISST annual SST}

## construct df for the GAM smoothing of mean SST
had_gam_df <- gather(had_gps, reef, sst, `1`:`5607`)
had_gam_df$date <- gsub("\\X", "", had_gam_df$date) # convert year to numeric and remove 'X'
had_gam_df <- had_gam_df %>% separate(date, c("year", "month","day"), convert = TRUE) # create year, month, and day column columns
had_gam_df$date <- as.Date(with(had_gam_df, paste(year, month, day, sep="-")), "%Y-%m-%d")


#### Data visualization per year across months #### 

# calculate summary statistic per month across years
HADI_year_avg <- had_full %>% 
  group_by(year) %>% 
  summarize(mean = mean(mean_temp, na.rm = TRUE),
            max = max(mean_temp, na.rm = TRUE),
            min = min(mean_temp, na.rm = TRUE),
            sd = sd(mean_temp, na.rm = TRUE),
            n = n()) %>% 
  mutate(se = sd / sqrt(n),
         lowerci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upperci = mean + qt(1 - (0.05 / 2), n - 1) * se,
         range = max - min)


## plot with smoothed trend
had_year_plot <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), text = element_text(size = 14)) +
  geom_hline(yintercept = mean(HADI_year_avg$mean), colour = "grey", linetype = 2) +
  geom_ribbon(data = HADI_year_avg, aes(x = year, ymin = lowerci, ymax = upperci), colour = "darkgrey", alpha = 0.3, linetype = 1) + 
  #geom_line(data = HADI_year_avg, aes(x = year, y = mean), colour = "black", group = 1, size = 1, alpha = 0.7) + # mean sst
  scale_x_continuous(name = "Year", breaks = c(1880, 1900, 1920, 1940, 1960, 1980, 2000, 2020)) +
  ylab('SST (°C)') +
  scale_y_continuous(breaks = seq(24, 30, 1)) +
  geom_line(data = HADI_year_avg, aes(x = year, y = max), colour = "#d6604d", group = 1, size = 0.5) + # max sst
  geom_line(data = HADI_year_avg, aes(x = year, y = min), colour = "#2166ac", group = 1, size = 0.5) + # min sst
  geom_smooth(data = had_gam_df, method = "gam", aes(x = year, y = sst), colour = "black", group = 1, size = 1, se = FALSE, formula = y ~ s(x, bs = "cs")) + # smoothed mean temp over all data
  coord_cartesian(xlim = c(1870, 2020)) 

```

```{r HadISST combined figure, fig.align='center', fig.width=6, fig.height=7.5}

# Saves the manuscript version of the HadISST historic SST
plot_grid(had_month_plot, had_year_plot, ncol = 1, labels = c("A", "B"), align = "hv") +
  ggsave("figures/Manuscript_figs/Fig2_SST_trends.pdf", width = 6, height = 7.5)

```

**Historic SST trend on Caribbean coral reefs (1871–2020).** Long-term SST records (HadISST) on Caribbean coral reefs depicting **A**) mean monthly SST each year (represented by line colour: blue to red) and **B**) GAM smoothed annual mean SST time (black line), annual maximum (red line), and annual minimum (blue line) SST. The grey dashed horizontal line denotes the overall mean SST for all sites over the entire period (27.3 °C) and the grey ribbon represents the 95% confidence interval around the true annual SST mean through time.

<br/>


### <span style="color: #0072B2;">**Figure 3**</span>

```{r HadISST ecoregion SST extraction - REMOTE, eval=FALSE, include=FALSE}

# ***NOTE***
####--- this portion of the script was run on a high performance computing cluster because of the long runtime 
####--- see separate R script (SSTExtraction_ReefGPS.R) for instructions (located in the code folder)
####--- the script can be put on a cluster with the HadISST netCDF (HadISST_sst.nc) to run


##########################################

### Extract SST using the ecoregion classifications and reef points

## Create an empty dataframe to populate while running the forloop with the following parameters:
eco_df <- data.frame('date' = integer(),
                     'sst' = integer(),
                     'ecoregion' = factor())

eco_sst_df <- data.frame('date' = integer(),
                         'sst' = integer(),
                         'ecoregion' = factor())


#### Run the forloop in parallel (if able):

## Read in the HadISST netCDF and clip based on Caribbean bounds
data_brick <- brick(had_data) # read SST data as rasterbrick
raster_brick <- crop(data_brick, extent(Xmin, Xmax, Ymin, Ymax)) # clip rasterbrick for Caribbean region only (see 'spatial subsetting' above)
layers <- raster_brick@data@nlayers

for(e in 1:length(eco_list)) {
 
  ## First, select reefs within specific ecoregion 
  ecoregion_name <- eco_list[[e]] # pull the name of ecoregion
  subset_ecoreg <- ecoregions2[ecoregions2@data$ECOREGION == ecoregion_name, ] # subset for only the desired ecoregion
  projection(subset_ecoreg) <- proj # modify the lat/lon projection to match reef points
  ecoregion_reefs <- reef_points[subset_ecoreg, ] # clip the reef points by the ecoregion bounds
  
  output <- foreach(i = 1:layers, .combine = "rbind", .packages = c("raster", "rgdal", "sf", "ncdf4", "tidyverse")) %dopar% {
    
    ## Read in the netCDF and clip based on Caribbean bounds
    data_brick <- brick("data/HadISST/HadISST_sst.nc") # read SST data as rasterbrick
    raster_brick <- crop(data_brick, extent(Xmin, Xmax, Ymin, Ymax)) # clip rasterbrick for Caribbean region only (see 'spatial subsetting' above)
    
    ## Then, extract SST from reef locations wihtin the current ecoregion and append to dataframe
    raster <- raster_brick[[i]] # select the layer of the rasterbrick 
    sst_vals <- raster::extract(raster, ecoregion_reefs) # extract SST at lat/lon of each reef
    # create dataframe of current layer (date) and extracted SST
    df <- data.frame('date' = raster@data@names,
                     'sst' = sst_vals)
    df$ecoregion <- rep(ecoregion_name, length(df$date)) # add a ecoregion name column
    eco_df <- rbind(eco_df, df) # adds current layer SST data to the full dataframe
    
    return(eco_df)
  }
  eco_sst_df <- rbind(eco_sst_df, output)
}



## Once completed, saves the resulting dataframe as both a .Rdata object and .csv
# *** Only uncomment these lines if you want to write over existing files ***
#save(eco_sst_df, file = "data/HadISST/Ecoregion_SST/HadISST_ecoregion_sst.Rdata")
#write.csv(eco_sst_df, file = "data/HadISST/Ecoregion_SST/HadISST_ecoregion_sst", row.names=FALSE)

```

```{r HadISST ecoregion monthly SST dataframe construction}

## Load the ecoregion extracted SST data (here using Rdata, but you can use csv if smaller)
load("data/HadISST/HadISST_ecoregion_sst.Rdata")

# Rename "Northern GoM" and "Northern GoM" to just GoM (these will be considered the same ecoregion from here on)
eco_sst_df$ecoregion <- gsub("Northern G", "G", eco_sst_df$ecoregion)
eco_sst_df$ecoregion <- gsub("Southern G", "G", eco_sst_df$ecoregion)

ecoregion_sst <- eco_sst_df %>% 
  group_by(ecoregion, date) %>% 
  summarize(mean_sst = mean(sst, na.rm = TRUE),
            max = max(sst, na.rm = TRUE),
            min = min(sst, na.rm = TRUE),
            n = n())


ecoregion_sst <- ecoregion_sst %>% separate(date, c("year", "month", "day"), convert = TRUE) # create year, month, and day column columns
ecoregion_sst$year <- as.numeric(gsub("\\X", "", ecoregion_sst$year)) # convert the year column to numerics and remove 'X'
ecoregion_sst$month2 <- factor(ecoregion_sst$month) # make a month as a factor column for plotting  

```

```{r HadISST ecoregion monthly SST per year, eval=FALSE, fig.align='center', fig.height=10, fig.width=9, include=FALSE}

#### Data visualization of monthly SST per year #### 

## create plots of monthly HadISST per year
# markdown plot
ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), axis.text.x = element_text(angle = 70, hjust = 1)) +
  geom_line(data = ecoregion_sst, aes(x = month, y = mean_sst, group = year, colour = year), alpha = 0.6, size = 0.2) +
  scale_colour_gradient2("Year", low = "#2166ac", mid = "#d1e5f0", high = "#d6604d", midpoint = mean(ecoregion_sst$year), space="Lab") +
  scale_x_continuous(name = "", breaks = c(1,2,3,4,5,6,7,8,9,10,11,12), labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  scale_y_continuous(breaks = seq(20, 32, 2)) +
  ylab('SST (°C)') +
  facet_wrap(~ecoregion, ncol = 2) +
  guides(colour = guide_colorbar(frame.colour = "black", ticks.colour = "black"))

```

```{r HadISST ecoregion annual SST, fig.height=9, fig.width=9}

## Calculate summary statistic per month across years
ecoregion_sst_annual <- ecoregion_sst %>% 
  group_by(ecoregion, year) %>% 
  summarize(mean = mean(mean_sst, na.rm = TRUE),
            max = max(mean_sst, na.rm = TRUE),
            min = min(mean_sst, na.rm = TRUE),
            range = max - min,
            sd = sd(mean_sst, na.rm = TRUE),
            n = n()) %>% 
  mutate(se = sd / sqrt(n),
         lowerci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upperci = mean + qt(1 - (0.05 / 2), n - 1) * se)

## Create overall mean SST per ecoregion for plotting dashed line
ecoregion_year_mean <- ecoregion_sst %>% 
  group_by(ecoregion) %>% 
  summarize(mean = round(mean(mean_sst, na.rm = TRUE), 2),
            range = round(mean((max(mean_sst, na.rm = TRUE) - min(mean_sst, na.rm = TRUE)), na.rm = TRUE), 2))

ecoregion_year_mean$lab <- paste(LETTERS[1:8])

#### Data visualization per year across months #### 

## Plot with smoothed trend
ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), axis.text.x = element_text(angle = 30, hjust = 0.8), text = element_text(size = 14)) +
  geom_segment(data = ecoregion_year_mean, aes(y = mean, yend = mean, x = 1870, xend = 2020), colour = "grey", linetype = 2) +
  geom_ribbon(data = ecoregion_sst_annual, aes(x = year, ymin = lowerci, ymax = upperci), colour = "darkgrey", alpha = 0.3, linetype = 1) + 
  scale_x_continuous(name = "", breaks = seq(1880, 2020, 20)) +
  ylab('SST (°C)') +
  geom_line(data = ecoregion_sst_annual, aes(x = year, y = max), colour = "#d6604d", group = 1, size = 0.5) +
  geom_line(data = ecoregion_sst_annual, aes(x = year, y = min), colour = "#2166ac", group = 1, size = 0.5) + 
  geom_smooth(data = ecoregion_sst_annual, method = "gam", aes(x = year, y = mean), colour = "black", group = 1, size = 0.5, se = FALSE, formula = y ~ s(x, bs = "cs")) + # smoothed mean temp
  coord_cartesian(xlim = c(1870, 2020)) +
  scale_y_continuous(breaks = seq(20, 32, 2), limits = c(20, 32)) +
  facet_wrap(~ ecoregion, ncol = 2, scales = "free_y") +
  geom_text(data = ecoregion_year_mean, aes(x = -Inf, y = Inf, label = lab, group = ecoregion), fontface = "bold", size = 4.5, hjust = -0.5, vjust = 1.4) +
  ggsave("figures/Manuscript_figs/Fig3_Ecoregion_SST_trends.pdf", width = 7, height = 8)

```

**Historic SST trends on coral reefs within ecoregions (1871–2019).** Long-term SST records (HadISST) on Caribbean coral reefs separated by ecoregion depicting GAM smoothed annual mean SST time (black line), annual maximum (red line), and annual minimum (blue line) SST. The grey dashed horizontal line denotes the mean SST over the entire period and the grey ribbon represents the 95% confidence interval around the true annual SST mean for each ecoregion.

<br/>


### Figure S2

```{r Pathfinder raster and times}

## read in the Pathfinder raster
path_data <- pathfinder_combined_monthly_data_path
data_brick2 <- brick(path_data) # read SST data as rasterbrick
data_brick2 <- crop(data_brick2,extent(Xmin,Xmax,Ymin,Ymax)) # clip rasterbrick for Caribbean region only (see 'spatial subsetting' above)


ncdf_file <- nc_open(path_data)
#ncdf_file

nc_lats <- ncdf_file$dim$lat$vals #other way of looking into the data structure; extracting lats
nc_longs <- ncdf_file$dim$lon$vals #longs here
nc_longs[nc_longs>180] <- nc_longs[nc_longs>180]-360

#time handling
raw_times <- ncdf_file$dim$time$vals # times are seconds after the "start of time"
startoftime <- ncdf_file$dim$time$units # here's what they consider the start of time
nc_times<-as.POSIXct(raw_times, tz="UTC", gsub("seconds since ","", startoftime))
nc_times <- nc_times+3600*3 # to center readings at noon - just cosmetics
nc_close(ncdf_file) # close the file

# temporal subsetting  
start_date2 <- "1981-09-15"
end_date2 <- "2019-12-16"
#end_date2 <- "1982-08-16" # using this one for just a subset of 12 months
start_date2 <- as.POSIXct(start_date2, tz="UTC")
end_date2 <- as.POSIXct(end_date2, tz="UTC")
    
# restrict temporally
Tstart_index2 <- min(which(difftime(nc_times,start_date2)>0))
Tend_index2 <- min(which(difftime(nc_times,end_date2)>0))
data_brick2 <- data_brick2[[Tstart_index2:Tend_index2]]
time <- nc_times[Tstart_index2:Tend_index2]

# save time component
save(time, nc_times, file = "data/Pathfinder/Pathfinder_times.Rdata")# save the time variables for IDs for running scripts remotely

```

```{r Pathfinder raster edit and slope calculation - REMOTE, eval=FALSE, include=FALSE}

# ***NOTE***
####--- this portion of the script was run on a high performance computing cluster because of the long runtime 
####--- see separate R script (Pathfinder_RasterCalculations.R) for instructions (located in the code folder)
####--- the script can be put on a cluster with the Pathfinder netCDF (pathfinder_combined_monthly_data.nc) to run


##########################################

## Remove NAs and convert to C
KtoC <- function(x){round(x-273.15,2)} # converts the raster from Kelvin to Celsius
NAset <- function(x){x[x < -54]<-NA; return(x)} # Pathfinder uses the value -54 (K) to denote missing data and this function replaces that with NA
data_brick2 <- calc(data_brick2, NAset) # apply the NA function written above to raster
data_brick2b <- calc(data_brick2, KtoC) # apply the Kelvin function written above to raster


## Calculate slope of temperature change across time for each pixel
# simple lm function applied to the raster brick to calculate the slope of temperature change over timescale of raster
lm.fun2 <- function(x, time){
  if(is.na(x)){NA} # this removes any NA values from the slope calculations 
  else{
    lm(x~time)$coefficients[2]*60*60*24 # slope here is degrees/day
  }}

raster_slope2 <- calc(data_brick2, function(x)lm.fun2(x, time = time)) # function applied here


## Create dataframe from calculated slope raster in C per decade for plotting
raster_slope2 <- raster_slope2*(365.25*10) # this is currently C per decade
path_slope <- as.data.frame(as(raster_slope2, "SpatialPixelsDataFrame")) # convert to a raster to da$
colnames(path_slope) <- c("sst", "x", "y")


## Save the edited netCDF file and the slope raster/dataframe
## Only uncomment these lines if you want to write over existing files!! 

#writeRaster(data_brick2b, filename = "data/Pathfinder/Pathfinder_monthly_edit.nc", format="CDF", overwrite=FALSE)
#save(path_slope, file = "data/Pathfinder/Pathfinder_slope.Rdata")
#write.csv(path_slope, file = "data/Pathfinder/Pathfinder_slope.csv", row.names=FALSE)

# --------- *Plotting occurs down at the bottom of the markdown file* --------- #

```

```{r Pathfinder pvalue map df creation - REMOTE, eval=FALSE, include=FALSE}

# ***NOTE***
####--- This does take a while (~10 minutes), however it can be run locally.
####--- see separate R script (Pathfinder_RasterCalculations.R) for instructions (located in the code folder)
####--- the script can be put on a cluster with the Pathfinder netCDF (pathfinder_combined_monthly_data.nc) to run


##########################################

## load in the edited pathfinder raster
path_brick <- brick(Pathfinder_monthly_edit_path)

## function to pull the p value per pixel of the simple lm applied to the full HadISST raster brick 
pval.fun <- function(x, time)
{
  if(any(is.na(x))){NA}
  else{
    summary(lm(x ~ time))$coefficients[4] # pulls the p value
  }
}
raster_pval <- calc(data_brick, function(x)pval.fun(x, time = time))

## function to pull the p value per pixel of the simple lm applied to the full Pathfinder raster brick 
raster_pval2 <- calc(path_brick, function(x)pval.fun(x, time = time))

# convert the raster to a df for plotting
path_pavl <- as.data.frame(as(raster_pval2, "SpatialPixelsDataFrame")) # convert to a raster to dataframe
colnames(path_pavl) <- c("pval", "x", "y")
path_pavl$bins <- cut(path_pavl$pval, breaks = bins, labels = bin_names) # bins the pvales into 6 unique bins for plotting

## save the dataframe to save time
write.csv(path_pavl, file = "data/Pathfinder/Pathfinder_pval.csv", row.names=FALSE)

# --------- *Plotting occurs down at the bottom of the markdown file* --------- #

```

```{r Pathfinder GPS extraction, eval=FALSE, include=FALSE}

# This takes a while to run, however, it was run locally.

load("data/Pathfinder/Pathfinder_times.Rdata")

## read in the Pathfinder raster
path_data <- "data/Pathfinder/PathfinderSST_monthly_edit.nc"
data_brick2 <- brick(path_data) # read SST data as rasterbrick
data_brick2 <- crop(data_brick2,extent(Xmin,Xmax,Ymin,Ymax)) # clip rasterbrick for Caribbean region only (see 'spatial subsetting' above)

reef_gps <- read.csv("data/ReefData/CaribbeanReef_points.csv") # read in FULL reef locations csv

## convert sampling points to spatial points and applying projection to points
LongLat <- cbind(reef_gps$lon, reef_gps$lat) 
ReefPTS <- SpatialPoints(LongLat)
proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84"
projection(ReefPTS) <- proj


## extract data from the sst brick with the sampling points shifted
path_sst <- raster::extract(data_brick2, ReefPTS)   # path_sst <- raster::extract(data_brick2, ReefPTS_shifted)


## convert the data matrix to a final dataframe
path_sst_xts <- xts(t(path_sst), time) # t is used to transpose the matrix because xts assumes dates are along the lines, not columns
path_sst_df <- data.frame(date=index(t(path_sst_xts)), coredata(t(path_sst_xts))) # convert xts object to dataframe
path_sst_df$reef <- factor(seq(1, length(path_sst_df$date), 1)) # ID the columns to identify unique reefs
path_sst_df$date <- NULL # remove this column since it is repetitive 
path_sst_long <- gather(path_sst_df, date, sst, X1981.09.15.17.34.40:X2019.12.16.11.45.55) # convert from wide to long format

## update the dataframe for saving
path_gps <- path_sst_long %>% separate(date, c("year", "month", "day", "hour", "min", "sec")) # create year, month, and day column columns
path_gps$date <- paste(path_gps$year, path_gps$month, path_gps$day, sep = ".") # creates a full date column (XYear.Month.Day)
path_gps <- path_gps[, -c(2:7)] # remove columns for hour, minute, and seconds (we won't use these)
path_gps <- spread(path_gps, reef, sst)

## save the final dataframe (.csv or Rdata)
#write.csv(path_gps, file = "data/Pathfinder/Pathfinder_reefs_SST_update.csv") # to save file as .csv
saveRDS(path_gps, file = "data/Pathfinder/Pathfinder_reefs_SST_update.rds") # to save file as Rdata (I saved as Rdata since it is large)

```

```{r Pathfinder monthly SST dataframe construction}

path_gps <- readRDS("data/Pathfinder/Pathfinder_reefs_SST.rds")

## create a new dataframe of simple stats per time
path_full <- data.frame("date" = path_gps[,1],
                        "mean_temp" = rowMeans(path_gps[,-1], na.rm = TRUE),
                        "max_temp" = apply(path_gps[,-1], 1, FUN=max, na.rm = TRUE),
                        "min_temp" = apply(path_gps[,-1], 1, FUN=min, na.rm = TRUE),
                        "n" = length(path_gps[,-1]))

path_full <- path_full %>% separate(date, c("year", "month","day"), convert = TRUE) # create year, month, and day column columns
path_full$year <- as.numeric(gsub("\\X", "", path_full$year)) # convert the year column to numerics and remove 'X'
path_full$month2 <- factor(path_full$month) # make a month as a factor column for plotting

```

```{r Pathfinder monthly SST per year, fig.align='center'}

#### Data visualization of monthly SST per year #### 

# calculate summary statistic per month across years
Path_month_avg <- path_full %>% 
  group_by(month) %>% 
  summarize(mean = mean(mean_temp, na.rm = TRUE),
            sd = sd(mean_temp, na.rm = TRUE),
            n = n()) %>% 
  mutate(se = sd / sqrt(n),
         lowerci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upperci = mean + qt(1 - (0.05 / 2), n - 1) * se)

# make year integer for better plotting
path_full$year <- as.integer(path_full$year)

## create plots of monthly Pathfinder per year
# markdown plot
path_month_plot <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 14)) +
  theme(legend.position = c(0.065, 0.85), legend.key.size = unit(0.3, "cm"), legend.background = element_rect(fill = NA), plot.title = element_text(size = 12)) + # legend theme
  guides(colour = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
  geom_line(data = path_full, aes(x = month, y = mean_temp, group = year, colour = year), alpha = 0.6, size = 0.4) +
  scale_colour_gradient2("", low = "#2166ac", mid = "#d1e5f0", high = "#d6604d", midpoint = mean(path_full$year), space="Lab") +
  scale_x_continuous(name = "Month", breaks = c(1,2,3,4,5,6,7,8,9,10,11,12), labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  scale_y_continuous(breaks = seq(24, 30, 1)) +
  ylab('SST (°C)') +
  ggtitle("Pathfinder (1981 - 2019)") + 
  theme(plot.title = element_text(size = 13))

```
  
```{r Pathfinder annual SST}

## construct df for the GAM smoothing of mean SST
path_gam_df <- gather(path_gps, reef, sst, `1`:`5607`)
path_gam_df$date <- gsub("\\X", "", path_gam_df$date) # convert year to numeric and remove 'X'
path_gam_df <- path_gam_df %>% separate(date, c("year", "month","day"), convert = TRUE) # create year, month, and day column columns
path_gam_df <- subset(path_gam_df, year != "2020") # exclude 2020 data


#### Data visualization per year across months #### 

# calculate summary statistic per month across years
Path_year_avg <- path_full %>% 
  group_by(year) %>% 
  summarize(mean = mean(mean_temp, na.rm = TRUE),
            max = max(mean_temp, na.rm = TRUE),
            min = min(mean_temp, na.rm = TRUE),
            sd = sd(mean_temp, na.rm = TRUE),
            n = n()) %>% 
  mutate(se = sd / sqrt(n),
         lowerci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upperci = mean + qt(1 - (0.05 / 2), n - 1) * se,
         range = max - min)


## plot with smoothed trend
path_year_plot <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), text = element_text(size = 14)) +
  #geom_vline(xintercept = 1981, colour = "lightgrey") +
  geom_hline(yintercept = mean(Path_year_avg[-1,]$mean), colour = "grey", linetype = 2) +
  geom_ribbon(data = Path_year_avg[-1,], aes(x = year, ymin = lowerci, ymax = upperci), colour = "darkgrey", alpha = 0.3, linetype = 1) + 
  #geom_line(data = Path_year_avg[-1,], aes(x = year, y = mean), colour = "black", group = 1, size = 1, alpha = 0.7) + # mean sst
  scale_x_continuous(name = "Year") +
  ylab('SST (°C)') +
  scale_y_continuous(breaks = seq(24, 30, 1)) +
  geom_line(data = Path_year_avg[-1,], aes(x = year, y = max), colour = "#d6604d", group = 1, size = 0.5) + # max sst
  geom_line(data = Path_year_avg[-1,], aes(x = year, y = min), colour = "#2166ac", group = 1, size = 0.5) + # min sst
  geom_smooth(data = path_gam_df, method = "gam", aes(x = year, y = sst), colour = "black", group = 1, size = 1, se = FALSE, formula = y ~ s(x, bs = "cs")) + # smoothed mean temp over all data
  coord_cartesian(xlim = c(1981, 2019))

```

```{r Pathfinder combined figure, fig.align='center', fig.width=6, fig.height=7.5}

# Saves the manuscript version of the Pathfinder historic SST
plot_grid(path_month_plot, path_year_plot, ncol = 1, labels = c("A", "B"), align = "hv") +
  ggsave("figures/Supplemental_figs/FigS2_Pathfinder_SST_trends.pdf", width = 6, height = 7.5)

```

Historic SST records (1981-2019; Pathfinder) on Caribbean coral reefs depicting **A**) mean monthly SST each year (represented by line colour: blue to red) and **B**) GAM smoothed annual mean SST time (black line), annual maximum (red line), and annual minimum (blue line) SST. The grey dashed horizontal line denotes the overall mean SST for all sites over the entire period (27.23 °C) and the grey ribbon represents the 95% confidence interval around the true annual SST mean through time. 

<br/>


### Figure S3

```{r combined annual SST plot}

## plot with smoothed trend
combo_plot <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  # add the HadISST dataset
  geom_hline(yintercept = mean(HADI_year_avg$mean), colour = "darkgrey", linetype = 2) +
  geom_ribbon(data = HADI_year_avg, aes(x = year, ymin = lowerci, ymax = upperci), colour = "darkgrey", alpha = 0.05, linetype = 1) + 
  geom_line(data = HADI_year_avg, aes(x = year, y = max), colour = "#d6604d", group = 1, size = 0.5, alpha = 0.3) +
  geom_line(data = HADI_year_avg, aes(x = year, y = min), colour = "#2166ac", group = 1, size = 0.5, alpha = 0.3) + 
  geom_line(data = HADI_year_avg, aes(x = year, y = mean), colour = "darkgrey", group = 1, size = 0.5, alpha = 0.3) + 
  geom_smooth(data = HADI_year_avg, method = "gam", aes(x = year, y = mean), formula = y ~ s(x, bs = "cs"), colour = "darkgrey", group = 1, size = 1, se = FALSE, alpha = 0.3) +
  # add the Pathfinder dataset on top
  geom_hline(yintercept = mean(Path_year_avg[-1,]$mean), colour = "black", linetype = 2) +
  geom_ribbon(data = Path_year_avg[-1,], aes(x = year, ymin = lowerci, ymax = upperci), colour = "darkgrey", alpha = 0.3, linetype = 1) +
  scale_x_continuous(name = "Year", breaks = c(1880, 1900, 1920, 1940, 1960, 1980, 2000, 2020)) +
  ylab('SST (°C)') +
  geom_line(data = Path_year_avg[-1,], aes(x = year, y = max), colour = "#d6604d", group = 1, size = 0.5) +
  geom_line(data = Path_year_avg[-1,], aes(x = year, y = min), colour = "#2166ac", group = 1, size = 0.5) + 
  geom_line(data = Path_year_avg[-1,], aes(x = year, y = mean), colour = "black", group = 1, size = 0.5) + 
  geom_smooth(data = Path_year_avg[-1,], method = "gam", aes(x = year, y = mean), formula = y ~ s(x, bs = "cs"), colour = "black", group = 1, size = 1, se = FALSE) +
  ggtitle("Caribbean coral reefs historic SST trends") +
  scale_y_continuous(breaks = seq(24, 30, 1)) +
  coord_cartesian(xlim = c(1870, 2020)) +
  ggsave("figures/Supplemental_figs/FigS3_Both_SST_trends.pdf", width = 6, height = 4)

ggplotly(combo_plot)

```

Comparison of HadISST (1871-2020) and Pathfinder (1981-2019) SST recorded on Caribbean coral reef locations. The high-resolution Pathfinder is represented as darker data over the long-term HadISST. Both datasets are represented by GAM smoothed annual mean SST time (solid line), annual maximum (red line), and annual minimum (blue line) SST. The dashed horizontal line denotes the overall mean SST for all sites over the entire period and the grey ribbon represents the 95% confidence interval around the true annual SST mean through time. 

<br/>


### Figure S4
*Still need to update this figure*

<br/>


### HadISST GIF

```{r HadISST GIF, fig.align='center'}

## create version of the plot for creating a GIF of HadISST data
had_plot_gif <- ggplot() +
  theme_bw() +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill=NA, size=1), axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(size = 20)) +
  guides(colour = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
  geom_line(data = had_full, aes(x = month, y = mean_temp, group = year, colour = year), size = 1) +
  scale_colour_gradient2("Year", low = "#2166ac", mid = "#d1e5f0", high = "#d6604d", midpoint = mean(had_full$year), space="Lab") +
  #geom_ribbon(data = HADI_month_avg, aes(x = month, ymin = lowerci, ymax = upperci), colour = "black", alpha = 0.6, linetype = 2) + 
  #geom_line(data = HADI_month_avg, aes(x = month, y = mean), colour = "black", group = 1, size = 1.5) +
  scale_x_continuous(name = "Month", breaks = c(1,2,3,4,5,6,7,8,9,10,11,12), labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  ylab('SST (°C)')

## below is code to transform the HadISST monthly SST plot into a GIF (not shown in the Rmarkdown file, but saved in the figures folder)
# SST is plotted one year at a time in order in the GIF

had_gif <- had_plot_gif + 
  transition_time(year) +
  labs(title = "Year: {frame_time} (HadISST)") +
  shadow_mark(alpha = 0.1, size = 0.5)

animate(had_gif, fps = 10, nframes = 300, end_pause = 2, width = 650, height = 400)
anim_save("figures/Supplemental_figs/HadISST.gif")

```


Long-term SST records (HadISST) on Caribbean coral reefs depicting mean monthly SST each year (represented by line colour: blue to red).

<br/>


## Caribbean Basin Maps and Marine Heatwaves {.tabset}

### <span style="color: #0072B2;">**Figure 4**</span>

```{r Pathfinder SST slope map}

### Pathfinder SST slope map

## Clip Pathfinder slopes by MHW bounds
load("data/Pathfinder/Pathfinder_slope.Rdata") # load Rdata of the modified Pathfinder raster
path_slope <- path_slope[c(2,3,1)] # just reordering the dataframe
coordinates(path_slope) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(path_slope) <- proj # modify projection to match Pathfinder
path_clipped <- path_slope[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
path_slope_clipped <- as.data.frame(path_clipped) # convert to dataframe

## Plot Pathfinder slope map
path_map <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = path_slope_clipped, aes(x = x, y = y, fill = sst), alpha=0.8, interpolate = FALSE) + 
  scale_fill_distiller("ºC per decade", palette = "RdYlBu", limits = c(-0.3, 0.4)) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  theme(panel.background = element_rect(fill = "grey95")) +
  labs(x = "", y = "") +
  ggtitle(paste0("Caribbean warming rate (1981-2019)"))

# save a version for the markdown visual
path_map_markdown <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom") +
  geom_raster(data = path_slope_clipped, aes(x = x, y = y, fill = sst), alpha=0.8, interpolate = FALSE) + 
  scale_fill_distiller("ºC per decade", palette = "RdYlBu", limits = c(-0.3, 0.4)) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  theme(panel.background = element_rect(fill = "grey95")) +
  labs(x = "", y = "") +
  ggtitle(paste0("Caribbean warming rate (1981-2019)"))



## Clip Pathfinder pvalues by MHW bounds
path_pavl <- read.csv("data/Pathfinder/Pathfinder_pval.csv", header = TRUE) # read in the saved .csv of the dataframe
path_pavl <- path_pavl[c(2,3,1, 4)] # just reordering the dataframe
coordinates(path_pavl) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(path_pavl) <- proj # modify projection to match Pathfinder
pathP_clipped <- path_pavl[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
pathP_slope_clipped <- as.data.frame(pathP_clipped) # convert to dataframe

## Plot Pathfinder significance map
path_pval_map <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = pathP_slope_clipped, aes(x = x, y = y, fill = bins), alpha=0.8, interpolate = FALSE) + 
  scale_fill_manual("p-value", values = "#045a8d") +
  #scale_fill_brewer("p-value", palette = "PuBu", direction = 1) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  theme(panel.background = element_rect(fill = "grey95")) +
  labs(x = "", y = "") +
  ggtitle("Significance of Caribbean warming (1981-2019)")

# save a version for the markdown visual
path_pval_markdown <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom") +
  geom_raster(data = pathP_slope_clipped, aes(x = x, y = y, fill = bins), alpha=0.8, interpolate = FALSE) + 
  scale_fill_manual("p-value", values = "#045a8d") +
  #scale_fill_brewer("p-value", palette = "PuBu", direction = 1) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  theme(panel.background = element_rect(fill = "grey95")) +
  labs(x = "", y = "") +
  ggtitle("SST warming (1981-2019)")

```

```{r OISST MHW frequency map}

### MHW slope map
MHW_trends <- read_rds(MHW_path) # load in trend data
MHW_trends$lon <- MHW_trends$lon- 360 # convert lon back into -180 to 180

# add a second pval column- for the breaks used in the map_p plot below
MHW_trends <- MHW_trends %>% 
  mutate(pvalb = cut(p, breaks = c(0, 0.001, 0.01, 0.05, .10, .50, 1)))


## MHW slopes map
map_slope <- ggplot(MHW_trends, aes(x = lon, y = lat)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  #geom_rect(size = 0.2, fill = NA, aes(xmin = lon - 0.1, xmax = lon + 0.1, ymin = lat - 0.1, ymax = lat + 0.1, colour = pval)) +
  geom_raster(aes(fill = slope), interpolate = FALSE, alpha = 0.9) +
  scale_fill_distiller("MHW per year", palette = "RdYlBu", limits = c((min(MHW_trends$slope)-0.04), max(MHW_trends$slope))) +
  #scale_color_brewer("MHW per year", palette = "RdYlBu", name = "p-value", guide = FALSE) +
  geom_polygon(data = map_base, aes(group = group), colour = "#575757", fill = "grey81") +
  theme(panel.background = element_rect(fill = "grey95")) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  ggtitle("Change in marine heat wave frequency (1981-2018)")

# markdown version of above
mhw_map_markdown <- ggplot(MHW_trends, aes(x = lon, y = lat)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom") +
  #geom_rect(size = 0.2, fill = NA, aes(xmin = lon - 0.1, xmax = lon + 0.1, ymin = lat - 0.1, ymax = lat + 0.1, colour = pval)) +
  geom_raster(aes(fill = slope), interpolate = FALSE, alpha = 0.9) +
  scale_fill_distiller("MHW per year", palette = "RdYlBu", limits = c((min(MHW_trends$slope)-0.04), max(MHW_trends$slope))) +
  #scale_color_brewer("MHW per year", palette = "RdYlBu", name = "p-value", guide = FALSE) +
  geom_polygon(data = map_base, aes(group = group), colour = "#575757", fill = "grey81") +
  theme(panel.background = element_rect(fill = "grey95")) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  ggtitle("Change in MHW frequency (1981-2018)")



## MHW significance map
map_p <- ggplot(MHW_trends, aes(x = lon, y = lat)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(aes(fill = pvalb), interpolate = FALSE) +
  scale_fill_brewer("p-value", palette = "PuBu", direction = -1, breaks = c("(0,0.001]", "(0.001,0.01]", "(0.01,0.05]", "(0.05,0.1]", "(0.1,0.5]", "(0.5,1]")) +
  geom_polygon(data = map_base, aes(group = group), colour = "#575757", fill = "grey81") +
  theme(panel.background = element_rect(fill = "grey95")) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  ggtitle("Significance of change in heat wave frequency (1981-2018)")

# markdown version of above
mhw_pval_markdown <- ggplot(MHW_trends, aes(x = lon, y = lat)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom") +
  geom_raster(aes(fill = pvalb), interpolate = FALSE) +
  scale_fill_brewer("p-value", palette = "PuBu", direction = -1, breaks = c("(0,0.001]", "(0.001,0.01]", "(0.01,0.05]", "(0.05,0.1]", "(0.1,0.5]", "(0.5,1]")) +
  geom_polygon(data = map_base, aes(group = group), colour = "#575757", fill = "grey81") +
  theme(panel.background = element_rect(fill = "grey95")) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  ggtitle("Change in MHW frequency (1981-2018)")

```

```{r combined SST and MHW maps}

### Combined plots
Figure4 <- cowplot::plot_grid(path_map, map_slope, labels = c("A", "B"), label_size = 18) +
  ggsave("figures/Manuscript_figs/Fig4_SST_MHW_map.pdf", width = 14.5, height = 5)

# markdown version
cowplot::plot_grid(path_map_markdown, mhw_map_markdown, labels = c("A", "B"))

```

**Warming patterns throughout the Caribbean Sea.** Increasing warming events across the Caribbean depicted through **A**) rate of SST change (°C per decade) from 1981 to 2019 (Pathfinder; mean slope `r round(mean(path_slope_clipped$sst), 2)` ± `r round(sd(path_slope_clipped$sst), 3)` °C per decade) and **B**) increasing marine heatwave events (slope of counts per year). Grey ocean area was not included in these analyses.

---

<br/>


### <span style="color: #0072B2;">**Figure 5**</span>
*Need to add MHW metrics here*

---

<br/>


### Figure S5

```{r combined SST and MHW p value maps}

## Combined p values:
FigureS5 <- cowplot::plot_grid(path_pval_map, map_p, align = "hv", labels = c("A", "B")) +
  ggsave("figures/Supplemental_figs/FigS5_SST_MHW_significance_map.pdf", width = 14.5, height = 5)

# markdown version of plot
cowplot::plot_grid(path_pval_markdown, mhw_pval_markdown, align = "hv", labels = c("A", "B")) 

```

Significance of **A**) rate of SST change (°C per decade) and **B**) number of marine heatwave events per year across the Caribbean depicted in **Figure 4**. Grey ocean area was not included in these analyses.

---

<br/>


### Figure S6

```{r HadISST full SST map, message=FALSE, warning=FALSE}

## Clip HadISST slopes by MHW bounds
had_slope_b <- had_slope[c(2,3,1)] # just reordering the dataframe
coordinates(had_slope_b) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(had_slope_b) <- proj # modify projection to match Pathfinder
had_clipped <- had_slope_b[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
had_slope_clipped <- as.data.frame(had_clipped) # convert to dataframe

## Plot the map
had_map <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = had_slope_clipped, aes(x = x, y = y, fill = sst), alpha=0.8, interpolate = FALSE) + 
  scale_fill_distiller("ºC per decade", palette = "RdYlBu", limits = c(min(had_slope_clipped$sst), max(had_slope_clipped$sst))) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "grey95")) +  
  ggtitle(paste0("Caribbean warming rate (", format.Date(start_date, "%Y"), "-", format.Date(end_date, "%Y"), ")"))


```

```{r HadISST full SST pval map, message=FALSE, warning=FALSE}

## Clip Pathfinder slopes by MHW bounds
had_pavl_b <- had_pavl[c(2,3,1, 4)] # just reordering the dataframe
coordinates(had_pavl_b) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(had_pavl_b) <- proj # modify projection to match Pathfinder
had_p_clipped <- had_pavl_b[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
had_pavl_clipped <- as.data.frame(had_p_clipped) # convert to dataframe


## Plot the map
had_pval_map <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = had_pavl_clipped, aes(x = x, y = y, fill = bins), alpha=0.8, interpolate = FALSE) +
  scale_fill_manual("p-value", values = "#045a8d") +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "grey95")) +  
  ggtitle("SST slope significance")

```

```{r HadISST full SST map combo, message=FALSE, warning=FALSE, fig.align='center', fig.width=12, fig.height=6}

plot_grid(had_map, had_pval_map, labels = c("A", "B")) +
  ggsave("figures/Supplemental_figs/FigS6_HadISST_map.pdf", width = 14, height = 5)

```

**A**) Rate of SST change (°C per decade) over the duration of the HadISST database across the Caribbean from 1871 to 2020 (mean slope `r round(mean(had_slope_clipped$sst), 2)` ± `r round(sd(had_slope_clipped$sst), 3)` °C per decade) and **B**) significance of rate of SST change. Grey ocean area was not included in these analyses.

---

<br/>


### Figure S7

```{r HadISST SST map subset (1981-2019), message=FALSE, warning=FALSE}

## Clip HadISST slopes by MHW bounds
had_slope_sub_b <- had_slope_sub[c(2,3,1)] # just reordering the dataframe
coordinates(had_slope_sub_b) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(had_slope_sub_b) <- proj # modify projection to match Pathfinder
had_s_slope_clipped <- had_slope_sub_b[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
had_slope_sub_clipped <- as.data.frame(had_s_slope_clipped) # convert to dataframe

## Plot the map
had_map_sub <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = had_slope_sub_clipped, aes(x = x, y = y, fill = sst), alpha=0.8, interpolate = FALSE) + 
  scale_fill_distiller("ºC per decade", palette = "RdYlBu", limits = c(min(had_slope_sub_clipped$sst), max(had_slope_sub_clipped$sst))) +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  guides(fill = guide_colorbar(frame.colour = "black", ticks.colour = "black", barwidth = 12)) +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "grey95")) +  
  ggtitle(paste0("Caribbean warming rate (", format.Date(start_dateb, "%Y"), "-", format.Date(end_date, "%Y"), ")"))

```

```{r HadISST SST pval map subset (1981-2019), message=FALSE, warning=FALSE}

## Clip Pathfinder slopes by MHW bounds
had_pval_sub_b <- had_pval_sub[c(2,3,1, 4)] # just reordering the dataframe
coordinates(had_pval_sub_b) <- c("x", "y") # transform lat/lon to SpatialPointsDataFrame
projection(had_pval_sub_b) <- proj # modify projection to match Pathfinder
had_p_sub_clipped <- had_pval_sub_b[bound_SP_df, ] # clip the Pathfinder SST slope by the new MHW bounds
had_pval_sub_b_clipped <- as.data.frame(had_p_sub_clipped) # convert to dataframe


## Plot the map
had_pval_map_sub <- ggplot() +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background=element_blank(), strip.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1), legend.position = "bottom", text = element_text(size = 15), title = element_text(size = 13)) +
  geom_raster(data = had_pval_sub_b_clipped, aes(x = x, y = y, fill = bins), alpha=0.8, interpolate = FALSE) +
  scale_fill_manual("p-value", values = "#045a8d") +
  geom_polygon(data = map_base, aes(group = group, x = lon, y = lat), colour = "#575757", fill = "grey81") +
  coord_sf(xlim = c(-100,-55), ylim = c(7.65,33), expand = FALSE) +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "grey95")) +  
  ggtitle("SST slope significance")

```

```{r HadISST SST map subset combo, message=FALSE, warning=FALSE, fig.align='center', fig.width=12, fig.height=6}

plot_grid(had_map_sub, had_pval_map_sub, labels = c("A", "B")) +
  ggsave("figures/Supplemental_figs/FigS7_HadISST_subset_map.pdf", width = 14, height = 5)

```

**A**) Rate of SST change (°C per decade) across the Caribbean from 1981 to 2020 (HadISST; mean slope `r round(mean(had_slope_sub_clipped$sst), 2)` ± `r round(sd(had_slope_sub_clipped$sst), 3)` °C per decade) and **B**) significance of rate of SST change. Grey ocean area was not included in these analyses.

---

<br/>


## Supplemental Tables {.tabset}

### Table S1
**Table S1.** Marine heatwave properties examined in this study

| Metric | Units | Description  |
|--------|:------|:-------------|
| Frequency | Number of events | Number of discrete mhw events |
| Total MHW days | days | The total number of days a location experienced a MHW per year |
| Duration | days | The number of days between the start and end date of each distinct MHW event |
| Peak intensity | °C | The maximum temperature, above the seasonal varying climatological mean, reached during the MHW event |  
| Onset rate | °C/day | “Rate of temperature change between onset date and date of peak intensity” |
| Return time | days | The number of days elapsed since a previous mhw event in that location |


### Table S2
**Table S2.** Calculated warming rates from both HadISST and Pathfinder databases for different temporal ranges.

```{r HadISST warming rates for table}
#### Basin values
### Caribbean Basin (1871-2020)

## Rate of warming
had_basin_full_rate <- round(mean(had_slope_clipped$sst), 2)

## Total C
had_basin_full_value <- round(((had_basin_full_rate/10) * 150), 2) # multiply number of years by the yearly rate for total warming

### Caribbean Basin (1990-2020)

## Rate of warming
had_basin_sub_rate <- round(mean(had_slope_sub_clipped$sst), 2)

## Total C
had_basin_sub_value <- round(((had_basin_sub_rate/10) * 31), 2) # multiply number of years by the yearly rate for total warming


#### Reef-specific values
### Caribbean Reefs (1871-2020)

## Rate of warming
had_sst <- had_gam_df # renaming the GAM dataframe for analyses below
had_reef_full_lm <- lm(sst ~ year, data = had_sst) # regression of temperature and year to calculate rate of warming
had_reef_full_rate <- round(had_reef_full_lm$coefficients[2]*10, 2) # extract the slope of rate and multiply it by 10 for decade rate

## Total C
had_reef_full_value <- round(((had_reef_full_rate/10) * 150), 2) # multiply number of years by the yearly rate for total warming


### Caribbean Reefs (1990-2020)

## Rate of warming
had_sst_sub <- subset(had_sst, year >= 1990) # subset data for 1990-2020 only
had_reef_sub_lm <- lm(sst ~ year, data = had_sst_sub) # regression of temperature and year to calculate rate of warming (since 1990)
had_reef_sub_rate <- round(had_reef_sub_lm$coefficients[2]*10, 2) # extract the slope of rate and multiply it by 10 for decade rate

## Total C
had_reef_sub_value <- round(((had_reef_sub_rate/10) * 31), 2) # multiply number of years by the yearly rate for total warming

```

```{r Pathfinder warming rates for table}

#### Basin values
### Caribbean Basin (1871-2020)

## Rate of warming
path_basin_full_rate <- round(mean(path_slope_clipped$sst), 2)

## Total C
path_basin_full_value <- round(((path_basin_full_rate/10) * 39), 2) # multiply number of years by the yearly rate for total warming


#### Reef-specific values
### Caribbean Reefs (1981-2020)

## Rate of warming
path_sst <- path_gam_df # renaming the GAM dataframe for analyses below
path_reef_full_lm <- lm(sst ~ year, data = path_sst) # regression of temperature and year to calculate rate of warming
path_reef_full_rate <- round(path_reef_full_lm$coefficients[2]*10, 2) # extract the slope of rate and multiply it by 10 for decade rate

## Total C
path_reef_full_value <- round(((path_reef_full_rate/10) * 38), 2) # multiply number of years by the yearly rate for total warming


### Caribbean Reefs (1990-2020)

## Rate of warming
path_sst_sub <- subset(path_sst, year >= 1990) # subset data for 1990-2020 only
path_reef_sub_lm <- lm(sst ~ year, data = path_sst_sub) # regression of temperature and year to calculate rate of warming (since 1990)
path_reef_sub_rate <- round(path_reef_sub_lm$coefficients[2]*10, 2) # extract the slope of rate and multiply it by 10 for decade rate

## Total C
path_reef_sub_value <- round(((path_reef_sub_rate/10) * 31), 2) # multiply number of years by the yearly rate for total warming

```

```{r table S2 compilation and visualization}

## Compile dataframe from above calculated values/rates

table_S2 <- data.frame("Temperature parameter" = c("Caribbean Basin (°C per decade)", "Caribbean Basin (total °C for period)", "Caribbean Reefs (°C per decade)", "Caribbean Reefs (total °C for period)"),
                       "HadISST (1871-2020)" = c(had_basin_full_rate, had_basin_full_value, had_reef_full_rate, had_reef_full_value),
                       "HadISST (1990-2020)" = c(had_basin_sub_rate, had_basin_sub_value, had_reef_sub_rate, had_reef_sub_value),
                       "Pathfinder (1981-2019)" = c(path_basin_full_rate, path_basin_full_value, path_reef_full_rate, path_reef_full_value),
                       "Pathfinder (1990-2019)" = c("NA", "NA", path_reef_sub_rate, path_reef_sub_value)
)


# View the completed 
kable(table_S2, booktabs = TRUE, row.names = FALSE, col.names = c("Temperature parameter", "HadISST (1871-2020)", "HadISST (1990-2020)", "Pathfinder (1981-2019)", "Pathfinder (1990-2019)")) %>% 
  kable_styling(font_size = 12, full_width = FALSE)

```


### Table S3
**Table S3.** Mean ocean warming rate (°C per decade; with 95% confidence interval) and total increase in temperature (°C) on coral reefs within each Caribbean ecoregion since the noted year of inflection point. The inflection point for each ecoregion was identified as the year in which annual warming rates noticeably increased based on the GAM smoothed annual means (**see Figures 3 and S4**). Years indicated with an asterisk (*) represent inflection points identified for ecoregions that occurred before the Pathfinder dataset record, therefore those rates and total warming values were calculated from the beginning of the record in 1981.  

```{r Pathfinder ecoregion SST dataframe construction}

## Load the ecoregion extracted SST data (here using Rdata, but you can use csv if smaller)
eco_sst_path_df <- readRDS("data/Pathfinder/Pathfinder_ecoregion_sst.rds")
load("data/Pathfinder/Pathfinder_times.Rdata") # load the Pathfinder R data containing 'times'

dates <- data.frame(date = paste("X", rep(1:460, 1), sep = ""),
                    date3 = time,
                    date2 = time)



# Rename "Northern GoM" and "Northern GoM" to just GoM (these will be considered the same ecoregion from here on)
eco_sst_path_df$ecoregion <- gsub("Northern G", "G", eco_sst_path_df$ecoregion)
eco_sst_path_df$ecoregion <- gsub("Southern G", "G", eco_sst_path_df$ecoregion)

eco_sst_path_df <- left_join(eco_sst_path_df, dates, by = c("date" = "date")) # left-join dataframes by date
eco_sst_path_df <- eco_sst_path_df %>% separate(date2, c("year", "month", "day"), convert = TRUE) # create year, month, and day column columns
eco_sst_path_df$month2 <- factor(eco_sst_path_df$month) # make a month as a factor column for plotting  

```

```{r Pathfinder extract warming rate from ecoregions}

region_names <- levels(factor(eco_sst_path_df$ecoregion)) # pull list of the different ecoregions
# make a dataframe of ecoregion and the year of warming increase
ecoregion_warm_year <- data.frame(ecoregion = region_names,
                                  warm_year = c(1986, 1985, 1981, 1986, 1981, 1981, 1981, 1993))

# join the warming year to the SST dataframe
eco_sst_path_df <- eco_sst_path_df %>% 
  left_join(ecoregion_warm_year, by = c("ecoregion" = "ecoregion"))

# create a dataframe to fill in the forloop
ecoregion_slope <- data.frame(ecoregion = character(),
                              warm_year = numeric(),
                              slope = numeric(),
                              lw_CI = numeric(),
                              up_CI = numeric())

# forloop that will calculate linear rate of warming per region along with confidence intervals
for (region in 1:length(region_names)) {
  
  ecoregion_subset <- subset(eco_sst_path_df, ecoregion == ecoregion_warm_year[region, 1]) # subset for selected ecoregion
  ecoregion_subset <- subset(ecoregion_subset, year >= ecoregion_warm_year[region, 2]) # subset for year of warming
  ecoregion_lm <- lm(sst ~ year, data = ecoregion_subset) # perform regression of sst through time
  slope <- round(ecoregion_lm[["coefficients"]][["year"]], 3) # extract the slope of SST over time
  lw_CI <- round(confint(ecoregion_lm, level = 0.95)[2,1], 4) # extract the lower CI 
  up_CI <- round(confint(ecoregion_lm, level = 0.95)[2,2], 4) # extract the upper CI 
  
  # add data to dataframe and convert rates to per decade 
  slope_df <- data.frame(ecoregion = ecoregion_warm_year[region, 1],
                         warm_year = ecoregion_warm_year[region, 2],
                         slope = (slope * 10),
                         lw_CI = (lw_CI * 10),
                         up_CI = (up_CI * 10))
  
  ecoregion_slope <- rbind(ecoregion_slope, slope_df) # add to the existing dataframe
}

# calculate the total amount warmed per ecoregion
ecoregion_slope$total_warm <- round(((ecoregion_slope$slope / 10) * (2019 - ecoregion_slope$warm_year)), 2)

```

```{r table S3 visualization}

## Compile dataframe from ecoregion rates
kable(ecoregion_slope, booktabs = TRUE, row.names = FALSE, col.names = c("Ecoregion", "Inflection point", "Mean rate (°C per decade)", "Lower confidence interval", "Upper confidence interval", "Warming extent (°C)")) %>% 
  kable_styling(font_size = 12, full_width = FALSE)

```


## Session information {.tabset}

All code was written by [Colleen B. Bove](https://colleenbove.science), feel free to contact with questions.

Session information from the last run date on `r date`:

```{r session info}

sessionInfo()

```